An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-07 main.tex
more about Parikh automata
New graphs, figure placement, etc

To illustrate the key features of \Calculus, we will informally present our
reasoning. We begin by considering the following set of string and other
constraints, which we want to solve for all possible values of $\MonoidElement$,
a value in some commutative monoid and the string $s$, where
$\varphi(\MonoidElement)$ is some set of constraints on $\MonoidElement$:
\begin{equation*}
  \AndComp{}{\begin{aligned}
  & \mathtt{ac^*a|b(bb)^*c} \text{ accepts } s\\
  & \mathtt{.^*c.^*} \text{ accepts } s \\
  & \MonoidElement = f(s) \\
  & \varphi(\MonoidElement)
  \end{aligned}}
 \end{equation*}

@@

\begin{figure}[t]
  \begin{minipage}[b]{0.70\linewidth}
  \centering 
  \includegraphics[scale=0.9]{aca_or_bc}
  \caption{An automaton recognising the regular expression
    $\mathtt{ac^*a|b(bb)^*c}$. \Fudge{this is actually not the right
      regular expression. the automaton also accepts words in which a
      b is followed by a.}}\label{fig:aca-or-bc}

    \includegraphics[scale=0.9]{something_c_something}
    \caption{An automaton recognising the regular expression
    $\mathtt{.^*c.^*}$.}\label{fig:something-c-something}
  \end{minipage}
  \Description[Two deterministic finite automata, one above the other]{NFAs 
    recognising the languages ac^*a|b(bb)^*c and ac^*a|b(bb)^*c respectively.
    The first one has 4 states and the second one 2.}
  \end{figure}

@@

\begin{figure}[t]
  \centering 
    \includegraphics[width=0.50\linewidth]{aca_or_bc_simplified}
    \caption{$\AcaOrBc{}$ with its symbolic transition counters after
    simplification of their equations.}\label{fig:example-simplify-1}
    \Description[The automaton $\AcaOrBc{}$ labelled with variables counting the number of times each transition is taken, after having simplified the equations.]{The starting state shows an initial flow of 1, with two outgoing transitions to $A$ ($1-l_c'$ on letter a, and $l_b' + l_c'$ on b respectively). $A$ has a flow of $l_c$, and its outgoing transition into the final state, $F$ is the same as the incoming one from $S$, the initial state. The $B$ transition has an outgoing transition labeled $c/ l_c'$ to the final state $F$, and one backwards transition $b / l_b'$ to the starting state.}
  \end{figure}

@@

\begin{figure}[t]
  \centering 
    \includegraphics[width=0.4\linewidth]{something_c_something_length_simplified}
    \caption{$\SomethingCSomething{}$ after using linear reasoning over lengths to replace a variable.}\label{fig:example-length-reduced}
    \Description[The automaton $\SomethingCSomething{}$ after reasoning over lengths.]{The same automaton as before for $\SomethingCSomething{}$, but with the transition from start to final annotated with a constant number 1, and the leftmost self-loop transition replaced with the equation $1 + l_c + 2l_b' -
    r_\Sigma'$.}
  \end{figure}

@@

\begin{figure}[t]
  \centering 
    \includegraphics[width=0.75\linewidth]{aca_or_bc_length_split_1}
    \caption{$\AcaOrBc{}$ at the split where $\FromLabelTo{S}{a}{A}$ is not
    used.}\label{fig:aca-or-bc-length-split-1}
    \Description[The automaton \AcaOrBc{} where we have removed the transition
    from $S$ to $A$.]{ The state $A$ hangs free without incoming transitions,
    and only the lower part of the automaton remains.}
  \end{figure}

@@

  \begin{figure}[t]
    \centering 
      \includegraphics[width=0.5\linewidth]{aca_or_bc_length_split_2}
      \caption{$\AcaOrBc{}$ at the split where $\FromLabelTo{S}{a}{A}$ \emph{is} used.}\label{fig:aca-or-bc-length-split-2}
      \Description[$\AcaOrBc{}$ with the transition from $B$ to $F$ removed.]{$\AcaOrBc{}$ when we keep the transition from $S$ to $A$. This removes the transition between $B$ and $F$, but otherwise maintains the automaton.}
    \end{figure}

@@

\begin{figure}[t]
  \centering 
    \includegraphics[width=0.5\linewidth]{length_split_product}
    \caption{The product of $\AcaOrBc$ and $\SomethingCSomething$ after
    case splitting and removal of dead states.}\label{fig:length-split-product}
    \Description[$\AcaOrBc{}$ with the transition from $B$ to $F$ removed.]{$\AcaOrBc{}$ with the transition from $B$ to $F$ removed. There now remains only one path to an accepting state.}
  \end{figure}

@@

  \begin{figure}[t]
    \centering 
      \includegraphics[width=0.7\linewidth]{aca_or_bc_simplified_3}
      \caption{$\AcaOrBc{}$ after further constraint propagation.}\label{fig:example-simplify-3}
      \Description[$\AcaOrBc$ without the lower section.]{$\AcaOrBc$ with only its upper section left and the $a$ transitions now set to static 1.}
    \end{figure}

@@

\begin{figure}[t]
\centering
    \includegraphics[width=0.75\linewidth]{aca_or_bc_times_something_c_something}
    \caption{$\AcaOrBc{} \times \SomethingCSomething{}$ under the constraint
    $\TransitionCount{a} > \TransitionCount{b}$. Note the fresh variables
    introduced for the transitions (for most transitions the same one, since
    they are all linearly dependent).}\label{fig:final-example} \Description[The
    final product, still a boring stick automaton with only one loop for an
    arbitrary number of c's.]{A product of the intermittent automata, which
    accepts \lstinline{ac.*ca}}
  \end{figure}

@@

\subsubsection{An Example}
We apply \eqref{eq:generate-parikh} to the automaton $\AcaOrBc{}$ of \cref{fig:aca-or-bc}. We then get the quantified Presburger formula seen in \eqref{eq:parikh-formula}. As before, we omit the implicit existential quantifiers for brevity, along with the constranints requiring all variables to be nonnegative. We also omit the $\land$-clauses between each line. In other words, in this example, only the variables $\LetterVar_a, \LetterVar_b, \LetterVar_c, \LetterVar_d, \LetterVar_e$, and $\LetterVar_f$ are free, and corresponds to the counts of the respective subscript letter.

\begin{equation}
  \begin{aligned}
  \ParikhMap(\Automaton):=\:
  &\LetterVar_{a} = \TransitionVar_{\FromLabelTo{0}{a}{1}} + \TransitionVar_{\FromLabelTo{1}{a}{3}} \land \LetterVar_{b} = \TransitionVar_{\FromLabelTo{0}{b}{2}} + \TransitionVar_{\FromLabelTo{2}{b}{0}} \land \LetterVar_{c} = \TransitionVar_{\FromLabelTo{1}{c}{1}} + \TransitionVar_{\FromLabelTo{2}{c}{3}}\\
  &\FinalStateVar_{3} > 0 \implies \StateVar_{3} > 0\\
  &\TransitionVar_{\FromLabelTo{0}{b}{2}} + \TransitionVar_{\FromLabelTo{0}{a}{1}} - \TransitionVar_{\FromLabelTo{2}{b}{0}} = 1 \land  \TransitionVar_{\FromLabelTo{1}{a}{3}} - \TransitionVar_{\FromLabelTo{0}{a}{1}} = 0\\
  & \TransitionVar_{\FromLabelTo{2}{b}{0}} + \TransitionVar_{\FromLabelTo{2}{c}{3}} - \TransitionVar_{\FromLabelTo{0}{b}{2}} = 0 \land \FinalStateVar_{3}  - \TransitionVar_{\FromLabelTo{1}{a}{3}} - \TransitionVar_{\FromLabelTo{2}{c}{3}} = 0\\
  &\TransitionVar_{\FromLabelTo{0}{a}{1}} > 0 \implies \StateVar_{1} > 0 \land \TransitionVar_{\FromLabelTo{0}{b}{2}} > 0 \implies \StateVar_{2} > 0\\
  &\TransitionVar_{\FromLabelTo{1}{c}{1}} > 0 \implies \StateVar_{1} > 0 \land \TransitionVar_{\FromLabelTo{1}{a}{3}} > 0 \implies \StateVar_{3} > 0\\
  &\TransitionVar_{\FromLabelTo{2}{b}{0}} > 0 \implies \StateVar_{0} > 0 \land \TransitionVar_{\FromLabelTo{2}{c}{3}} > 0 \implies \StateVar_{3} > 0\\
  &\StateVar_{0} > 0 \implies \left(\StateVar_{0} = \StateVar_{1} + 1 \land \TransitionVar_{\FromLabelTo{0}{a}{1}} \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{0} = \StateVar_{2} + 1 \land \TransitionVar_{\FromLabelTo{0}{b}{2}} \geq 1 \land \StateVar_{2} \geq 1\right)\\
  &\StateVar_{1} > 0 \implies \left(\StateVar_{1} = \StateVar_{1} + 1 \land \TransitionVar_{\FromLabelTo{1}{c}{1}} \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{1} = \StateVar_{3} + 1 \land \TransitionVar_{\FromLabelTo{1}{a}{3}} \geq 1 \land \StateVar_{3} \geq 1\right)\\
  &\StateVar_{2} > 0 \implies \left(\StateVar_{2} = \StateVar_{0} + 1 \land \TransitionVar_{\FromLabelTo{2}{b}{0}} \geq 1 \land \StateVar_{0} \geq 1\right) \lor \left(\StateVar_{2} = \StateVar_{3} + 1 \land \TransitionVar_{\FromLabelTo{2}{c}{3}} \geq 1 \land \StateVar_{3} \geq 1\right)\\
  &\StateVar_{3} > 0 \implies \FinalStateVar_{3} = 1
  \end{aligned}
  \label{eq:parikh-formula}
  \end{equation}

  We see here that the formula encodes the choice of path at the start of the
  automaton ($\TransitionVar_{\FromLabelTo{0}{b}{2}} +
  \TransitionVar_{\FromLabelTo{0}{a}{1}} -
  \TransitionVar_{\FromLabelTo{2}{b}{0}} = 1$), and the dependency between that
  choice and the following outgoing transition, e.g. in
  $\TransitionVar_{\FromLabelTo{1}{a}{3}} -
  \TransitionVar_{\FromLabelTo{0}{a}{1}} = 0$. Here we only have one choice of
  incoming flow; from $\FinalStateVar_3$, corresponding to the only accepting
  state.

Note that one of the $\lor$-clauses can be immediately removed since it contains
the contradiction $\StateVar_{1} = \StateVar_{1} + 1$, arising from cycles.The
unsatisfiability is by design, and ensures that two states in a cycle cannot
vouch for each other's connectedness.

Assume we are really interested in occurrences with the letter a, i.e. we solve
for $\LetterVar_a > 0$. This implies that $\StateVar_{1} > 0 \land \StateVar_{3}
> 0$ (essentially, that the state our transition of interest started in is
reachable), in turn implying that $\FinalStateVar_3 = 1$. Plugging in these
facts, we get:
\begin{equation}
  \begin{aligned}
  &\LetterVar_{a} = 2 \land \LetterVar_{b} = 0 \land \LetterVar_{c} = \TransitionVar_{\FromLabelTo{1}{c}{1}}\\
  & \StateVar_{3} > 0\\
  &\StateVar_{0} > 0 \implies \StateVar_{0} = \StateVar_{3} + 2\\
  \end{aligned}
  \label{eq:parikh-formula-calc}
  \end{equation}

We can then proceed to eliminate the remaining existentially quantified
variables to obtain the formula $\LetterVar_{a} = 2 \land \LetterVar_{b} = 0
\land \LetterVar_{c} \geq 0$ without too much effort.

  This example, in addition to showing how the encoding works, has hopefully
  also illustrated one of the key insights behind the design of \Calculus: the
  fact that even few additional constraints on the Parikh image can dramatically
  reduce the difficulty of computing it.

@@

Intuitively, this produces a simpler problem than computing the Parikh image, as
we now ignore not just the order of characters but also their differences.
This
means that the branching automaton of \cref{fig:len-branch} becomes
equivalent to the non-branching automaton of \cref{fig:len-branch-free}.
Note that the elimination of the choice of paths also effectively eliminates the
possibility of disconnecting one of the self-loops. This, in turn, means that we
can eliminate all of the connectedness implications of the quantified formula,
which would in turn simplify it (and the quantifier elimination of it)
considerably.

@@

\iffalse
An example of a Parikh automata version of $\AcaOrBc{}$ can be seen in
\cref{fig:parikh-automaton}. A corresponding Presburger version of the
constraint seen in \cref{sec:motivation} is $\psi(\Vec{v}) = v_1 > v_2$.

Generally, we can encode the problem as follows:
\begin{enumerate}
  \item The morphism ($\Map$) we would use is $\pi_{v}$
  \item The monoid operation is element-wise vector addition, which is trivially commutative.
  \item \Fudge{We would just add the constraints on the final values!?}
\end{enumerate}
\fi

@@

  \Fudge{// This notation might be too much but IT'S OBVIOUS DAMMIT}

    The attentive reader will notice that the \GetModel{} function is
    nonstandard in that it returns both the model and the automaton used to
    generate it. Since we implement algorithm~\ref{alg:find-image} inside a theorem
    prover under our control, we are able to glean such internal data structures. This should be considered fair game, since we are in practice implementing a specialised form of quantifier elimination, which is an internal affair.

    The performance of this approach is evaluated in
    \cref{sec:evaluation:finding-image}, where we benchmark our implementation
    of it.

@@

@@

In satisfaction mode, supported by all back-ends, \Catra{} tries to satisfy the
constraints expressed by the input file, reporting satisfiable with register
assignments or unsatisfiable much like traditional~SAT- or SMT solvers would.
Additionally, the baseline and our own back-end also support efficient
generation of the Presburger formula describing the constraints of the input
file using the method described in \cref{sec:finding-the-image}.

@@

\begin{table}[ht]
  \centering
  \input{graphs/solved_pivot_table.tex}
  \caption{The result of running the respective back-ends by instance
  satisifiability (satisifiable or unsatisifiable) with a timeout of
  \RuntimeTimeout. Instances solved by no backend within the timeout are omitted
  from the table. }\label{tab:solve-status}
\end{table}

\begin{figure}[ht]
  \includegraphics[width=0.75\linewidth]{graphs/\commit-by-solver.pdf}
  \caption{The division of statuses per backend.}
  \label{fig:solve-division}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.75\linewidth]{graphs/\commit-time-boxplot.pdf}
  \caption{The distribution of runtimes for solved instances per backend. Note that the number of instances solved differs between backends.}
  \label{fig:runtime-boxplot}
\end{figure}

@@

\begin{figure}[ht]
  \includegraphics[width=0.75\linewidth]{graphs/\commit-cactus.pdf}
  \caption{The number of instances solved as the time budget increases, simulated from one \RuntimeTimeout-timeout run.}
  \label{fig:cactus}
\end{figure}

\subsection{Finding a Presburger Formula}\label{sec:evaluation:finding-image}

@@

\begin{figure}[ht]
  \caption{The number of instances \Catra{} was able to find the Presburger form of the image for within a given number of seconds per backend.}
  \label{fig:cactus:image}
\end{figure}

>>>>

To illustrate the key features of \Calculus, we will informally present our
reasoning. We begin by considering the following set of string and other
constraints, which we want to solve for all possible values of $\MonoidElement$,
a value in some commutative monoid and the string $s$, where
$\varphi(\MonoidElement)$ is some set of constraints on $\MonoidElement$:
\begin{equation*}
  \AndComp{}{\begin{aligned}
  & \mathtt{/}\SndRegex{}\mathtt{/} \text{ accepts } s\\
  & \mathtt{/}\FstRegex{}\mathtt{/} \text{ accepts } s \\
  & \MonoidElement = f(s) \\
  & \varphi(\MonoidElement)
  \end{aligned}}
 \end{equation*}

@@

\begin{figure}[p]
  \centering 
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{aca_or_bc}
    \caption{An automaton recognising the regular expression
      $\FstRegex{}$.}\label{fig:aca-or-bc}
      \Description[An NFA recognising $\FstRegex{}$]{An NFA recognising $\FstRegex{}$. It has 4 states.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth} 
    \includegraphics[width=\textwidth]{something_c_something}
    \caption{An automaton recognising the regular expression
    $\SndRegex{}$.}\label{fig:something-c-something}
    \Description[An NFA recognising $\SndRegex{}$]{An NFA recognising $\FstRegex{}$. It has 2 states.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{aca_or_bc_simplified}
    \caption{$\AcaOrBc{}$ with its symbolic transition counters after
    simplification of their equations.}\label{fig:example-simplify-1}
    \Description[The automaton $\AcaOrBc{}$ labelled with variables counting the number of times each transition is taken, after having simplified the equations.]{The starting state shows an initial flow of 1, with two outgoing transitions to $A$ ($1-l_c'$ on letter a, and $l_b' + l_c'$ on b respectively). $A$ has a flow of $l_c$, and its outgoing transition into the final state, $F$ is the same as the incoming one from $S$, the initial state. The $B$ transition has an outgoing transition labeled $c/ l_c'$ to the final state $F$, and one backwards transition $b / l_b'$ to the starting state.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{something_c_something_length_simplified}
    \caption{$\SomethingCSomething{}$ after using linear reasoning over lengths to replace a variable.}\label{fig:example-length-reduced}
    \Description[The automaton $\SomethingCSomething{}$ after reasoning over lengths.]{The same automaton as before for $\SomethingCSomething{}$, but with the transition from start to final annotated with a constant number 1, and the leftmost self-loop transition replaced with the equation $1 + l_c + 2l_b' -
    r_\Sigma'$.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
      \includegraphics[width=\textwidth]{aca_or_bc_length_split_1}
      \caption{$\AcaOrBc{}$ at the split where $\FromLabelTo{S}{a}{A}$ is not
      used.}\label{fig:aca-or-bc-length-split-1}
      \Description[The automaton \AcaOrBc{} where we have removed the transition
      from $S$ to $A$.]{ The state $A$ hangs free without incoming transitions,
      and only the lower part of the automaton remains.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
      \includegraphics[width=\textwidth]{aca_or_bc_length_split_2}
      \caption{$\AcaOrBc{}$ at the split where $\FromLabelTo{S}{a}{A}$ \emph{is} used.}\label{fig:aca-or-bc-length-split-2}
      \Description[$\AcaOrBc{}$ with the transition from $B$ to $F$ removed.]{$\AcaOrBc{}$ when we keep the transition from $S$ to $A$. This removes the transition between $B$ and $F$, but otherwise maintains the automaton.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
      \includegraphics[width=\textwidth]{length_split_product}
      \caption{The product of $\AcaOrBc$ and $\SomethingCSomething$ after
      case splitting and removal of dead states.}\label{fig:length-split-product}
      \Description[$\AcaOrBc{}$ with the transition from $B$ to $F$ removed.]{$\AcaOrBc{}$ with the transition from $B$ to $F$ removed. There now remains only one path to an accepting state.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
      \includegraphics[width=\textwidth]{aca_or_bc_simplified_3}
      \caption{$\AcaOrBc{}$ after further constraint propagation.}\label{fig:example-simplify-3}
      \Description[$\AcaOrBc$ without the lower section.]{$\AcaOrBc$ with only its upper section left and the $a$ transitions now set to static 1.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.6\textwidth}
        \includegraphics[width=\textwidth]{aca_or_bc_times_something_c_something}
        \caption{$\AcaOrBc{} \times \SomethingCSomething{}$ under the constraint
        $\TransitionCount{a} > \TransitionCount{b}$.}\label{fig:final-example} \Description[The
        final product, still a boring stick automaton with only one loop for an
        arbitrary number of c's.]{A product of the intermittent automata, which
        accepts \lstinline{ac.*ca}}
  \end{subfigure}    
  \end{figure}

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

    The \GetModel{} function is nonstandard in that it returns both the model
    and its associated automaton. Since algorithm~\ref{alg:find-image} is
    essentially a form of quantifier elimination and therefore an internal affair
    to the theorem prover, this should be considered fair game.

    A preliminary implementation of this approach is available as part
    of~\Catra, but has been neither optimised nor fully tested.

@@

In satisfaction mode, supported by all back-ends, \Catra{} tries to satisfy the
constraints expressed by the input file, reporting satisfiable with register
assignments or unsatisfiable much like traditional~SAT- or SMT solvers would.
Additionally, the baseline and our own back-end also support generation of the
Presburger formula describing the constraints of the input file.

@@

\begin{table}
  \centering
  \input{graphs/solved_pivot_table.tex}
  \caption{The result of running the respective back-ends by instance
  satisifiability (satisifiable or unsatisifiable) with a timeout of
  \RuntimeTimeout. Instances solved by no backend within the timeout are omitted
  from the table. }\label{tab:solve-status}
\end{table}

\begin{figure}[t]
  \begin{minipage}[b]{0.49\linewidth}
    \centering
      \includegraphics[width=\textwidth]{graphs/\commit-by-solver.pdf}
      \caption{The division of statuses per backend.}
      \label{fig:solve-division}
      \Description[A bar chart showing three bars, one per backend, illustrating how many instances they could solve]{The bars are divided by satisfiable and unsatisfiable instances. Baseline could seemingly only solve unsatisfiable instances, lazy could solve a few satisfiable and mostly unsatisfiable, and nuxmv could solve about twice as many unsatisfiable as satisfiable instances, and slightly more in total than lazy.}
  \end{minipage}
  \begin{minipage}[b]{0.49\linewidth}
    \centering
      \includegraphics[width=\textwidth]{graphs/\commit-time-boxplot.pdf}
      \caption{The distribution of runtimes for solved instances per backend. Note that the number of instances solved differs between backends.}
      \label{fig:runtime-boxplot}
      \Description[A box plot showing the distribution of runtime over the three backends]{The middle box plot shows a tiny box centered around 0 seconds for the lazy backend, the rightmost box shows a bigger box between five seconds and 30 seconds for nuxmv, and the leftmost box shows a smaller box between 20 and 30 seconds for baseline. The whiskers of nuxmv span between 0 and the timeout, 60 seconds, while they are much tighter for lazy. On the other hand, lazy has  a lot of outliers.}
  \end{minipage}
  \vspace{0.5cm}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[width=0.75\textwidth]{graphs/\commit-cactus.pdf}
    \caption{The number of instances solved as the time budget increases, simulated from one \RuntimeTimeout-timeout run.}
    \label{fig:cactus}
    \Description[A cactus plot comparing the performance of nuxmv, lazy and baseline]{The line for baseline is strictly lower than the two others, and does not head upwards from zero until after 10 seconds of timeout, while the number of instances for nuxmv increases sharply until ca 10 seconds and then linearly after that. Lazy solves much more instances than nuxmv on shorter runtimes, but is overtaken by just before the 60-second timeout used in these experiments.}
  \end{minipage}
\end{figure}

@@

\iffalse
\subsection{Finding a Presburger Formula}\label{sec:evaluation:finding-image}

@@

\begin{figure}[ht]
  \caption{The number of instances \Catra{} was able to find the Presburger form of the image for within a given number of seconds per backend.}
  \label{fig:cactus:image}
\end{figure}
\fi
